Stability concerns in random matrices

### 2018-03-16 random_matrix_stability.Rmd
More simulations included
More thoughts on stability

I am specifically interested in the case in which $\boldsymbol{D}$ is known to be stable. Using random matrices containing element values that are real numbers uniformly distributed between $-4$ and $4$, I will show that rapid evolution is almost never stabilising and almost always destabilising when over six species and their traits $m$ are considered. To do this, I will use the following code, which finds a stable random matrix $\boldsymbol{D}$ and assigns random element values (possibly, but not necessarily stable) to $\boldsymbol{A}$, $\boldsymbol{B}$, and $\boldsymbol{C}$. I will do this for 20000 replicates of species numbers from $n = 2$ to $n = 8$.

@@

Here I will only consider $\boldsymbol{A}$ and focus on the stability of ecological interactions given variation in generation time. I will use the same technique of @Patel2018 for varying ecological and evolutionary time scales, but a bit differently. I assume that the response of a particular species' growth rate (rows in $\boldsymbol{A}$) is consistently affected by a value $\varepsilon_{i}$; higher values of $\varepsilon_{i}$ correspond to shorter absolute generation times and lower values correspond to longer generation times. The matrix below $\boldsymbol{A}$ for two species $i$ and $j$ with generation times $\varepsilon_{i}$ and $\varepsilon_{j}$, respectively.

$$\boldsymbol{A} = \begin{bmatrix}
    \varepsilon_{i}\frac{\partial N_{i}W_{i}}{\partial N_{i}}  & \varepsilon_{i}\frac{\partial N_{i}W_{i}}{\partial N_{j}}  \\
    \varepsilon_{j} \frac{\partial N_{j}W_{j}}{\partial N_{i}}  & \varepsilon_{j} \frac{\partial N_{j}W_{j}}{\partial N_{j}}  \\
\end{bmatrix}.$$

I will allow random values of $\varepsilon_{i}$ to vary from $1$ to $100$, modelling two orders of magnitude of generation times. I will compare the stability of two sets of random community matrices -- one in which all $\varepsilon = 1$ and another in which $\varepsilon$ varies.

```{r, echo = TRUE, eval = FALSE}
nn     <- 8;
A1_stt <- 0;
A2_stt <- 0;
iter   <- 100000;
while(iter > 0){
    A1_dat  <- runif(n = nn * nn, min = -4, max = 4);
    A1      <- matrix(data = A1_dat, nrow = nn, ncol = nn);
    A2_dat  <- runif(n = nn * nn, min = -4, max = 4);
    A2      <- matrix(data = A2_dat, nrow = nn, ncol = nn);
    epsil   <- runif(n = nn, min = 1, max = 100);
    eps_dat <- rep(x = epsil, times = nn);
    eps_mat <- matrix(data = epsil, nrow = nn, ncol = nn, byrow = FALSE);
    A2      <- A2 * eps_mat;
    A1_stb  <- max(Re(eigen(A1)$values)) < 0;
    A2_stb  <- max(Re(eigen(A2)$values)) < 0;
    if(A1_stb == TRUE){
      A1_stt <- A1_stt + 1;
    }
    if(A2_stb == TRUE){
      A2_stt <- A2_stt + 1;
    }
    iter    <- iter - 1;
}
```

The number of stable random matrices when $\varepsilon$ does not vary was $13$, and the number of stable random matrices when it does was $48$. To further investigate, it is worth looking at **how the amount of variation in generation time in a community affects stability**. Below is some code to look at stability in random communities across different magnitudes of variance in generation time for communities with eight species.

```{r, echo = TRUE, eval = FALSE}
res  <- NULL;
mags <- 10^(0:5);
for(i in mags){
    nn     <- 8;
    A1_stt <- 0;
    A2_stt <- 0;
    iter   <- 1000000;
    while(iter > 0){
        A1_dat  <- runif(n = nn * nn, min = -4, max = 4);
        A1      <- matrix(data = A1_dat, nrow = nn, ncol = nn);
        A2_dat  <- runif(n = nn * nn, min = -4, max = 4);
        A2      <- matrix(data = A2_dat, nrow = nn, ncol = nn);
        epsil   <- runif(n = nn, min = 1, max = i);
        eps_dat <- rep(x = epsil, times = nn);
        eps_mat <- matrix(data = epsil, nrow = nn, ncol = nn, byrow = FALSE);
        A2      <- A2 * eps_mat;
        A1_stb  <- max(Re(eigen(A1)$values)) < 0;
        A2_stb  <- max(Re(eigen(A2)$values)) < 0;
        if(A1_stb == TRUE){
          A1_stt <- A1_stt + 1;
        }
        if(A2_stb == TRUE){
          A2_stt <- A2_stt + 1;
        }
        iter    <- iter - 1;
    }
    res <- rbind(res, c(A1_stt, A2_stt));
    print(i);
}
```

@@

>>>>

I am specifically interested in the case in which $\boldsymbol{D}$ is known to be stable. Using random matrices containing element values that are real numbers uniformly distributed between $-4$ and $4$, I will show that rapid evolution is almost never stabilising and almost always destabilising when over six species and their traits $m$ are considered. To do this, I will use the following code, which finds a stable random matrix $\boldsymbol{D}$ and assigns random element values (possibly, but not necessarily stable) to $\boldsymbol{A}$, $\boldsymbol{B}$, and $\boldsymbol{C}$. I will do this for 100000 replicates of species numbers from $n = 2$ to $n = 8$.

@@

Here I will only consider $\boldsymbol{A}$ and focus on the stability of ecological interactions given variation in generation time. I will use the same technique of @Patel2018 for varying ecological and evolutionary time scales, but a bit differently. I assume that the response of a particular species' growth rate (rows in $\boldsymbol{A}$) is consistently affected by a value $\gamma_{i}$ (I am using $\gamma_{i}$ instead of $\varepsilon_{i}$ to avoid confusing generation time variance with ecological versus evolutionary rate difference). Higher values of $\gamma_{i}$ correspond to shorter absolute generation times and lower values correspond to longer generation times. The matrix below $\boldsymbol{A}$ for two species $i$ and $j$ with generation times $\gamma_{i}$ and $\gamma_{j}$, respectively.

$$\boldsymbol{A} = \begin{bmatrix}
    \gamma_{i}\frac{\partial N_{i}W_{i}}{\partial N_{i}}  & \gamma_{i}\frac{\partial N_{i}W_{i}}{\partial N_{j}}  \\
    \gamma_{j} \frac{\partial N_{j}W_{j}}{\partial N_{i}}  & \gamma_{j} \frac{\partial N_{j}W_{j}}{\partial N_{j}}  \\
\end{bmatrix}.$$

I will first allow random values of $\gamma_{i}$ to vary from $1$ to $100$, modelling two orders of magnitude of generation times. I will compare the stability of two sets of random community matrices -- one in which all $\gamma = 1$ and another in which $\gamma$ varies.

```{r, echo = TRUE, eval = FALSE}
nn     <- 8;
A1_stt <- 0;
A2_stt <- 0;
iter   <- 100000;
while(iter > 0){
    A1_dat  <- runif(n = nn * nn, min = -4, max = 4);
    A1      <- matrix(data = A1_dat, nrow = nn, ncol = nn);
    A2_dat  <- runif(n = nn * nn, min = -4, max = 4);
    A2      <- matrix(data = A2_dat, nrow = nn, ncol = nn);
    gamval  <- runif(n = nn, min = 1, max = 100);
    eps_dat <- rep(x = gamval, times = nn);
    eps_mat <- matrix(data = gamval, nrow = nn, ncol = nn, byrow = FALSE);
    A2      <- A2 * eps_mat;
    A1_stb  <- max(Re(eigen(A1)$values)) < 0;
    A2_stb  <- max(Re(eigen(A2)$values)) < 0;
    if(A1_stb == TRUE){
      A1_stt <- A1_stt + 1;
    }
    if(A2_stb == TRUE){
      A2_stt <- A2_stt + 1;
    }
    iter    <- iter - 1;
}
```

The number of stable random matrices when $\gamma$ does not vary was $13$, and the number of stable random matrices when it does was $48$. To further investigate, it is worth looking at **how the amount of variation in generation time (range of $\gamma$) in a community affects stability**. Below is some code to look at stability in random communities across different magnitudes of variance in generation time for communities with eight species.

```{r, echo = TRUE, eval = FALSE}
res  <- NULL;
mags <- 10^(0:5);
for(i in mags){
    nn     <- 8;
    A1_stt <- 0;
    A2_stt <- 0;
    iter   <- 10000000;
    while(iter > 0){
        A1_dat  <- runif(n = nn * nn, min = -4, max = 4);
        A1      <- matrix(data = A1_dat, nrow = nn, ncol = nn);
        A2_dat  <- runif(n = nn * nn, min = -4, max = 4);
        A2      <- matrix(data = A2_dat, nrow = nn, ncol = nn);
        gamval  <- runif(n = nn, min = 1, max = i);
        gam_dat <- rep(x = gamval, times = nn);
        gam_mat <- matrix(data = gam_dat, nrow = nn, ncol = nn, byrow = FALSE);
        A2      <- A2 * gam_mat;
        A1_stb  <- max(Re(eigen(A1)$values)) < 0;
        A2_stb  <- max(Re(eigen(A2)$values)) < 0;
        if(A1_stb == TRUE){
          A1_stt <- A1_stt + 1;
        }
        if(A2_stb == TRUE){
          A2_stt <- A2_stt + 1;
        }
        iter    <- iter - 1;
    }
    res <- rbind(res, c(A1_stt, A2_stt));
    print(i);
}
```

@@

Code to print the above is below.

```{r, eval = FALSE, echo = TRUE}
random_comm <- function(dm = 2, iter = 10000){
  stabilised_fastevo <- NULL;
  unstabled_fastevo  <- NULL;
  nn           <- dm * dm;
  tot_stabled  <- 0;
  tot_ustabled <- 0;
  tot_eco_stab <- 0;
  stabilised_fastevo <- NULL;
  unstabled_fastevo  <- NULL;
  while(iter > 0){
    epsil   <- runif(n = dm, min = 1, max = 1000);
    eps_dat <- rep(x = epsil, times = dm);
    eps_mat <- matrix(data = c(epsil, epsil), nrow = dm, ncol = dm, byrow = FALSE);
    stabilised   <- 0;
    unstabilised <- 0;
    Adat <- runif(n = nn, min = -4, max = 4);
    Bdat <- runif(n = nn, min = -4, max = 4);
    Cdat <- runif(n = nn, min = -4, max = 4);
    A    <- matrix(data = Adat, nrow = dm, ncol = dm) * eps_mat;
    B    <- matrix(data = Bdat, nrow = dm, ncol = dm) * eps_mat;
    C    <- matrix(data = Cdat, nrow = dm, ncol = dm) * eps_mat;
    egD  <- 1;
    while(egD >= 0){
      Ddat <- runif(n = nn, min = -4, max = 4);
      D    <- matrix(data = Ddat, nrow = dm, ncol = dm) * eps_mat;
      egD  <- max(Re(eigen(D)$values))
    }
    eco_stable       <- max(Re(eigen(A)$values)) < 0;
    slow_evo_maxE    <- max(Re(eigen(A + B %*% solve(D) %*% (-C))$values));
    stable_fast_evo  <- slow_evo_maxE < 0;
    if(eco_stable == FALSE & stable_fast_evo == TRUE){
      list_ele <- length(stabilised_fastevo$A) + 1;
      stabilised_fastevo$A[[list_ele]] <- A;
      stabilised_fastevo$B[[list_ele]] <- B;
      stabilised_fastevo$C[[list_ele]] <- C;
      stabilised_fastevo$D[[list_ele]] <- D;
      stabilised   <- 1;
    }
    if(eco_stable == TRUE & stable_fast_evo == FALSE){
      list_ele <- length(unstabled_fastevo$A) + 1;
      unstabled_fastevo$A[[list_ele]] <- A;
      unstabled_fastevo$B[[list_ele]] <- B;
      unstabled_fastevo$C[[list_ele]] <- C;
      unstabled_fastevo$D[[list_ele]] <- D;
      unstabilised   <- 1;
    }
    if(stabilised == 1){
      tot_stabled <- tot_stabled + 1;
    }
    if(unstabilised == 1){
      tot_ustabled <- tot_ustabled + 1;
    }
    if(eco_stable == TRUE){
      tot_eco_stab <- tot_eco_stab + 1;
    }
    iter <- iter - 1;
    if(iter %% 1000000 == 0){
        print(c(dm, iter));
    }
  }
  results <- list(ecologically_stable = tot_eco_stab, 
                  stabilised = tot_stabled, destabilised = tot_ustabled,
                  stabilised_all   = stabilised_fastevo, 
                  destabilised_all = unstabled_fastevo);
  return(results);
}

NN       <- 20;
res_tabl <- matrix(data = 0, nrow = (NN-1), ncol = 5);
its      <- 100000000;
for(i in 2:NN){
  res              <- random_comm(dm = i, iter = its); 
  res_tabl[i-1, 1] <- i;                               # Species and traits
  res_tabl[i-1, 2] <- res$ecologically_stable;         # Total stable
  res_tabl[i-1, 3] <- res$destabilised;                # Total destabilised
  res_tabl[i-1, 4] <- its - res_tabl[i-1, 2];          # Total unstable
  res_tabl[i-1, 5] <- res$stabilised;                  # Total stabilised
  print(i);
}
```
